State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-17 analysis/paper/paper.Rmd
updated all citation calls, corrected a few instances of Lower Granite dam (should be Lower Granite Dam), and re-knit all versions

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

# set default theme for ggplot2
# theme_set(theme_pubr(base_family = "Times New Roman",
#                      base_size = 12))
theme_set(theme_pubr())

@@

@@

Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g. @Bue1998]. For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @hess2014monitoring; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O. mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run (arriving at Lower Granite between March 1 and August 17) Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019BiOp; @NFSC2015; @NPCC2014]. Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.

@@

However, using window counts as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes. In-person window counts where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species). Counts are then expanded to provide an estimate for the entire hour [@USACE2015]. For the remainder of the year (November through March), fish passage is video-taped for 10 hours each day; fish counters then read the video tapes and submit daily fish counts. Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. @Hatch1994 conducted a study of the counting system at Lower Granite dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts. There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts for both Chinook Salmon and steelhead, accounted for by estimates of passage rates while the counting window is closed of 3.5% (Chinook Salmon) and 6.6% (steelhead). They also found some species misidentification issues, which could result in either under- or over-counting of a particular species. While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than the past decade.

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@royle2008hierarchical] which separates process variance (e.g. week to week variance in true escapment) from observation error variance (e.g. observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension). Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. We then applied this model to Chinook Salmon and steelhead returns at Lower Granite dam for spawn years 2010 - 2019. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.

@@

All captured fish were anesthetized, speciated, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery. The most widely usee marking of hatchery fish is the adipose fin clip, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without an adipose fin clip, and these are referred to as unclipped hatchery fish, or hatchery no-clip (HNC). For adipose-intact (unclipped) adults, which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019). Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tqagging [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed teh dam. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, quantifying observer error is impossible. Capturing and accounting for known sources of error is prudent to minimize management decision risk. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019]. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b]. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report]. Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

# color version
ts_col = plot_df %>%
  ggplot(aes(x = Start_Date)) +
  geom_ribbon(aes(ymin = trap_lowCI,
                  ymax = trap_uppCI,
                  fill = "Trap Est."),
              alpha = 0.4) +
  geom_ribbon(aes(ymin = `2.5%`,
                  ymax = `97.5%`,
                  fill = "Model"),
              alpha = 0.4) +
  geom_point(aes(y = win_cnt,
                 shape = 'Window Count',
                 color = 'Window Count'),
             size = pt_size) +
  geom_point(aes(y = trap_est,
                 shape = 'Trap Est.',
                 color = 'Trap Est.'),
             size = pt_size) +
  geom_line(aes(y = `50%`,
                color = 'Model')) +
  geom_point(aes(y = `50%`,
                 shape = 'Model',
                 color = 'Model'),
             size = pt_size) +
  scale_color_manual(values = c('Model' = 'black',
                                'Window Count' = 'blue',
                                'Trap Est.' = 'red'),
                     name = "Source") +
  scale_fill_manual(values = c('Model' = 'black',
                                'Window Count' = 'blue',
                                'Trap Est.' = 'red'),
                     guide = "none") +
  scale_shape_manual(values = c("Model" = 18,
                                "Window Count" = 1,
                                "Trap Est." = 8),
                     name = "Source") +
  theme_pubr() +
  theme(legend.position = 'bottom') +
  facet_wrap(~ plot_title,
             scales = "free",
             ncol = 2) +
  labs(x = 'Date',
       y = 'Total Escapement (x 10,000)')

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

# set default theme for ggplot2
theme_set(theme_pubr())

@@

if(doc.type %in% c("docx", "html")) {
  theme_set(theme_pubr(base_family = "Times New Roman"))
}

@@

Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g. @Bue1998]. For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @Hess2014; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@McElhany2000] and facilitate assessments of population viability, extinction risk [@NWFSC2015; @Williams2016] and climate change vulnerability [@Crozier2019].

Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O. mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015]. As a result, Snake River spring-summer run (arriving at Lower Granite between March 1 and August 17) Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019; @NWFSC2015; @NPCC2014]. Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.

@@

However, using window counts as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes. In-person window counts where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species). Counts are then expanded to provide an estimate for the entire hour [@USACE2015]. For the remainder of the year (November through March), fish passage is video-taped for 10 hours each day; fish counters then read the video tapes and submit daily fish counts. Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. @Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts. There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts for both Chinook Salmon and steelhead, accounted for by estimates of passage rates while the counting window is closed of 3.5% (Chinook Salmon) and 6.6% (steelhead). They also found some species misidentification issues, which could result in either under- or over-counting of a particular species. While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than the past decade.

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g. week to week variance in true escapment) from observation error variance (e.g. observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension). Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. We then applied this model to Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.

@@

All captured fish were anesthetized, speciated, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery. The most widely usee marking of hatchery fish is the adipose fin clip, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without an adipose fin clip, and these are referred to as unclipped hatchery fish, or hatchery no-clip (HNC). For adipose-intact (unclipped) adults, which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019). Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tqagging [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed teh dam. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, quantifying observer error is impossible. Capturing and accounting for known sources of error is prudent to minimize management decision risk. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019]. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b]. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020]. Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

# color version
ts_col = plot_df %>%
  ggplot(aes(x = Start_Date)) +
  geom_ribbon(aes(ymin = trap_lowCI,
                  ymax = trap_uppCI,
                  fill = "Trap Est."),
              alpha = 0.4) +
  geom_ribbon(aes(ymin = `2.5%`,
                  ymax = `97.5%`,
                  fill = "Model"),
              alpha = 0.4) +
  geom_point(aes(y = win_cnt,
                 shape = 'Window Count',
                 color = 'Window Count'),
             size = pt_size) +
  geom_point(aes(y = trap_est,
                 shape = 'Trap Est.',
                 color = 'Trap Est.'),
             size = pt_size) +
  geom_line(aes(y = `50%`,
                color = 'Model')) +
  geom_point(aes(y = `50%`,
                 shape = 'Model',
                 color = 'Model'),
             size = pt_size) +
  scale_color_manual(values = c('Model' = 'black',
                                'Window Count' = 'blue',
                                'Trap Est.' = 'red'),
                     name = "Source") +
  scale_fill_manual(values = c('Model' = 'black',
                                'Window Count' = 'blue',
                                'Trap Est.' = 'red'),
                     guide = "none") +
  scale_shape_manual(values = c("Model" = 18,
                                "Window Count" = 1,
                                "Trap Est." = 8),
                     name = "Source") +
  theme_pubr(legend = "bottom") +
  facet_wrap(~ plot_title,
             scales = "free",
             ncol = 2) +
  labs(x = 'Date',
       y = 'Total Escapement (x 10,000)')
